Exact expressions for double descent and implicit regularization \\ via surrogate random design

### 2022-11-23 design/continuous_nips/continuous.tex
Working compile

  \documentclass[thesis.tex]{subfiles}
%  \usepackage[round]{natbib}
%  \usepackage{fullpage}
%  \usepackage{amsmath}
%  \usepackage{amssymb}
%  \usepackage{amsthm}
%  \usepackage{mathabx}
%  \usepackage{color}
%  \usepackage{cancel}
%  \usepackage{wrapfig}
%  \usepackage{subfigure}
%  \usepackage{caption}
%  \usepackage{xfrac}
%  \usepackage{algorithm}
%  \usepackage{algorithmic}
%  \usepackage{mdframed}
%  \usepackage{hyperref}
%  \usepackage{xcolor}
%  \hypersetup{
%  	colorlinks,
%  	linkcolor={red!40!gray},
%  	citecolor={blue!40!gray},
%  	urlcolor={blue!70!gray}
%  }
%  \usepackage{cleveref}

@@

% \begin{abstract}
  Double descent refers to the phase transition that is exhibited by
  the generalization error of unregularized learning models when varying the ratio
  between the number of parameters and the number of training
  samples. The recent success of highly over-parameterized machine learning
  models such as deep neural networks has motivated a theoretical analysis of
  the double descent phenomenon in classical models such as linear
  regression which can also generalize well in the over-parameterized
  regime. We provide the first exact non-asymptotic
  expressions for double descent of the minimum norm linear
  estimator. Our approach involves constructing a special
  determinantal point process  which we call surrogate random
  design, to replace the standard i.i.d.~design of the training
  sample. This surrogate design admits exact expressions for the mean
  squared error of the estimator while preserving the key properties
  of the standard design. We also establish an exact implicit
  regularization result for over-parameterized training samples. In
  particular, we show that, for the surrogate design, the implicit bias
  of the unregularized minimum norm estimator precisely corresponds to
  solving a ridge-regularized least squares problem on the population
  distribution. In our analysis we introduce a new mathematical tool of
  independent interest: the class of random matrices for which
  determinant commutes with expectation.
% \end{abstract}

@@

%  To
% obtain our precise results, we use a somewhat non-standard random design, which
% we term surrogate random design (see Section~\ref{s:determinantal} for a
% detailed discussion), and which we expect will be of more general interest.
% Informally, the goal of a surrogate random design is to modify an original
% design to capture its main properties while being ``nicer'' in some useful way.
% In Theorem~\ref{t:asymptotic} and Appendix \ref{sec:asymp-conj-details} we show, both
% theoretically and empirically, that our surrogate design accurately preserves
% the key properties of the original design when the data distribution is a
% multivariate Gaussian.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}{0.49\textwidth}%
    \subcaption{Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
    numerical estimates even for non-isotropic
    features. Eigenvalue decay leads to a steeper
    descent curve in the under-determined regime ($n<d$).}
    \includegraphics[width=\textwidth]{../figs/descent-intro}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.49\textwidth}
    \subcaption{The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.}
    \includegraphics[width=\textwidth]{../figs/descent-shrinkage}
  \end{subfigure}
  \caption{Illustration of the main results for $d=100$ and
    $\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
    eigenvalues decaying exponentially and scaled so that
    $\tr(\Sigmab^{-1})=d$. We use our surrogate
    formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
    \ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
    lines), accompanied by the empirical estimates based on the standard
    i.i.d.~design (error bars are three times the standard error of the
    mean). We consider three different condition numbers $\kappa$ of
    $\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
    i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
    $\w^*=\frac1{\sqrt{d}}\one$.}
  %\vspace{-5mm}
  \label{f:intro}
\end{figure}

\noindent
Proof of Theorem \ref{t:mse} is given in Appendix \ref{a:mse-proof}.
For illustration, we plot the MSE expressions in Figure~\ref{f:intro}a,
comparing them with empirical estimates of the true MSE under the
i.i.d.~design for a multivariate Gaussian distribution
$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
$100$ and vary the number of samples $n$, observing a double descent
peak at $n=d$. We observe that our theory aligns well with
the empirical estimates, whereas
previously, no such theory was available except for special
cases such as $\Sigmab=\I$ (more details in Theorem~\ref{t:asymptotic}
and Section~\ref{sec:asymp-conj-details}). The plots
show that varying the spectral decay of $\Sigmab$ has a significant effect on the
shape of the curve in the under-determined regime. We use the
horizontal line to denote the MSE of the null estimator
$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
decay rapidly, then the Moore-Penrose estimator suffers less error
than the null estimator for some values of $n<d$, and the curve
exhibits a local optimum in this regime.

One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$.
This expression is precisely the classical notion of \emph{effective
  dimension} for ridge regression regularized with
$\lambda_n$~\citep{ridge-leverage-scores}, and it arises here even though there is
no explicit ridge regularization in the problem being considered in
Theorem~\ref{t:mse}.
The global solution to the ridge regression task (i.e., $\ell_2$-regularized
least squares) with parameter $\lambda$ is defined as:
\begin{align*}
  \argmin_\w \Big\{\E_{\mu,y}\big[\big(\x^\top\w-y(\x)\big)^2\big]
  + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
  \lambda\I)^{-1}\v_{\mu,y},\quad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
\end{align*}
When Assumption \ref{a:linear} holds, then
$\v_{\mu,y}=\Sigmab_\mu\w^*$, however ridge-regularized least squares
is well-defined for much more general response models.
Our second result makes a direct connection between the (expectation
of the) unregularized minimum norm solution on the sample
and the global ridge-regularized solution.
While the under-determined regime (i.e., $n<d$) is of primary interest to us,
for completeness we state this result for arbitrary values of $n$ and $d$.
Note that, just like the definition of regularized least squares, this
theorem applies more generally than Theorem~\ref{t:mse}, in that it
does \emph{not} require the responses to follow any linear model as in
Assumption~\ref{a:linear} (proof in Appendix~\ref{s:unbiased-proof}).
\begin{theorem}[Implicit regularization of Moore-Penrose estimator]
  \label{t:unbiased}
  For $\mu$ satisfying
  % \footnote{The proof of Theorem \ref{t:unbiased}
  %   can be easily extended to probability measures $\mu$ that do not
  %   satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
  %   assumption here to simplify the presentation.}
  Assumption~\ref{a:general-position} and
  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$,
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big] =
    \begin{cases}
      (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} & \text{for }n<d,     \\
      \Sigmab_\mu^{-1}\v_{\mu,y}                 & \text{for }n \ge d,
    \end{cases}
  \end{align*}
  where, as in Theorem \ref{t:mse}, $\lambda_n$ is such that the effective dimension
  $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
\end{theorem}

@@

\section{Related work}
\label{s:related-work}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.5\textwidth]{../figs/descent-model-log}
  \caption{
    Surrogate MSE as a function of $d/n$, with $n$
    fixed to $100$ and varying dimension $d$ and condition number
    $\kappa$, for signal-to-noise ratio $\mathrm{SNR}=1$.}
  \label{f:model}
\end{figure}

@@

The first non-trivial property of the surrogate design $S_\mu^n$ is
that the expected sample size is in fact always equal to $n$, which we
prove in Appendix \ref{appx: proof-of-l-size}.
\begin{lemma} \label{l:size}
  Let $\Xb\sim S_\mu^n$ for any $n>0$.
  Then, we have $\E[\#(\Xb)] = n$.
\end{lemma}

\noindent
Our general template for computing expectations under
a surrogate design $\Xb\sim\S_\mu^n$ is to use the following expressions based on the
i.i.d.~random design $\X\sim\mu^K$:
\begin{align}
  \E[F(\Xb)] & =\begin{cases}
    \frac{\E[\det(\X\X^\top)F(\X)]}{\E[\det(\X\X^\top)]}
    \quad K\sim\Poisson(\gamma_n) & \text{for }n<d, \\[2mm]
    \frac{\E[\det(\X)^2F(\X)]}{\E[\det(\X)^2]}\hspace{3mm}
    \quad K=d                     & \text{for }n=d, \\[2mm]
    \frac{\E[\det(\X^\top\X)F(\X)]}{\E[\det(\X^\top\X)]}
    \quad K\sim\Poisson(\gamma_n) & \text{for }n>d.
  \end{cases}\label{eq:cases}
\end{align}
These formulas follow from Definitions \ref{d:det} and
\ref{d:surrogate} because the determinants $\det(\X\X^\top)$ and
$\det(\X^\top\X)$ are non-zero precisely in the regimes $n\leq d$ and
$n\geq d$, respectively, which is why we can drop the restrictions on the
range of the Poisson distribution.
We compute the normalization constants
by introducing the concept of determinant preserving random matrices,
discussed in Section \ref{s:dp}.
% can be obtained using the following formulas: if $\X\sim\mu^K$ then
% \begin{align*}
% \text{(Lemma~\ref{l:normalization})}&&
% \E\big[\det(\X\X^\top)\big]  &= \ee^{-\gamma_n}\det(\I +
% \gamma_n\Sigmab_\mu)&&\text{for } K\sim\Poisson(\gamma_n),\ n<d,\\
%   \text{(Lemma~\ref{l:cb})}&& \E\big[\det(\X)^2\big] &=
%   d!\det(\Sigmab_\mu),&&\text{for } K=n=d,\\
%   \text{(Lemma~\ref{l:poisson})}&& \E\big[\det(\X^\top\X)\big]
% &=\det(\gamma_n\Sigmab_\mu),&&  \text{for }K\sim\Poisson(\gamma_n),\ n>d.
% \end{align*}
% \begin{remark}
% We will use $Z_\mu^n$ as a shorthand for the above normalization constants.
% \end{remark}
% For the case of $n=d$, the normalization constant can be found in the
% literature \cite{expected-generalized-variance,correcting-bias}:
% We prove Lemmas \ref{l:poisson} and \ref{l:normalization} in Section
% \ref{s:dp} by introducing the
% concept of determinant preserving random matrices.
% The lemmas play a
% crucial role in deriving a number of new expectation formulas for the
% under- and over-determined surrogate designs that we use to prove Theorems~\ref{t:mse}
% and~\ref{t:unbiased} in Section~\ref{s:expectations}.
% On the other hand, Lemma  \ref{l:cb} and the design $S_\mu^d$ can be found in the
% literature \citep{expected-generalized-variance,correcting-bias-journal},
% and we will rely on those known results in this case.
% Importantly, the $n=d$ case offers a continuous transition between the under-
% and over-determined regimes because the distribution $S_\mu^n$
% converges to $S_\mu^d$ when $n$ approaches $d$ from above and
% below.

\paragraph{Proof sketch of Theorem \ref{t:mse}}
% We now highlight some key expectation formulas for surrogate designs, which are used to
% derive the MSE expressions from Theorem \ref{t:mse}. 
We focus here on the under-determined regime (i.e., $n<d$),
highlighting the key new expectation formulas we develop to derive the
MSE expressions for surrogate designs. A standard decomposition of the MSE yields:
\begin{align}
  \MSE{\Xb^\dagger\ybb}
   & = \E\big[\|\Xb^\dagger(\Xb\w^*+\xib)-\w^*\|^2\big]
  =\sigma^2\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big] +
  \w^{*\top}\E\big[\I-\Xb^\dagger\Xb\big]\w^*.\label{eq:mse-derivation}
\end{align}
Thus, our task is to find closed form expressions for the two
expectations above. The latter, which is the expected projection onto
the complement of the row-span of $\Xb$, is proven in
Appendix \ref{s:unbiased-proof}.
\begin{lemma}\label{l:proj}
  If  $\Xb\sim S_\mu^n$ and $n<d$, then we have:
  $\E\big[\I-\Xb^\dagger\Xb\big] = (\gamma_n\Sigmab_\mu+\I)^{-1}$.
\end{lemma}

\noindent
No such expectation formula is known for i.i.d.~designs, except when
$\mu$ is an isotropic Gaussian. In Appendix \ref{s:unbiased-proof}, we
also prove a generalization of Lemma \ref{l:proj} which is then used
to establish our implicit regularization result
(Theorem~\ref{t:unbiased}). We next give an expectation formula
for the trace of the Moore-Penrose inverse of the covariance
matrix for a surrogate design (proof in Appendix \ref{a:mse-proof}).
\begin{lemma}\label{l:sqinv-all}
  If  $\Xb\sim S_\mu^n$ and $n<d$, then:
  $\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
    =\gamma_n\big(1-
    \det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big)$.
\end{lemma}

@@

\noindent
We next give a few simple examples to provide some intuition. First, note
that every $1\times 1$ random matrix is determinant preserving simply
because taking a determinant is an identity transfomation in one
dimension. Similarly, every fixed matrix is determinant preserving because
in this case taking the expectation is an identity
transformation. In all other cases, however, Definition \ref{d:main}
has to be verified more carefully. Further examples (positive and
negative) follow.
\begin{example}
  If $\A$ has i.i.d. Gaussian entries $a_{ij}\sim\Nc(0,1)$, then
  $\A$ is d.p.~because $\E[\det(\A)]=0$.
\end{example}

\noindent
In fact, it can be shown that all random matrices with independent entries
are determinant preserving. However, this is not a necessary condition.
\begin{example}\label{e:rank-1}
  Let $\A = s\,\Z$, where $\Z$ is fixed with $\rank(\Z) = r$, and $s$
  is a scalar random variable. Then for $|\Ic|=|\Jc|=r$ we have
  \begin{align*}
    \E\big[\det(s\,\Z_{\Ic,\Jc})\big] & = \E[s^r]\det(\Z_{\Ic,\Jc})
    =\det\Big(\big(\E[s^r]\big)^{\frac1r}\,\Z_{\Ic,\Jc}\Big),
  \end{align*}
  so if $r=1$ then $\A$ is determinant preserving, whereas if $r>1$
  and $\Var[s]>0$ then it is not.
\end{example}

>>>>

\documentclass[../../thesis.tex]{subfiles}

%  \usepackage[round]{natbib}
%  \usepackage{fullpage}
%  \usepackage{amsmath}
%  \usepackage{amssymb}
%  \usepackage{amsthm}
%  \usepackage{mathabx}
%  \usepackage{color}
%  \usepackage{cancel}
%  \usepackage{wrapfig}
%  \usepackage{subfigure}
%  \usepackage{caption}
%  \usepackage{xfrac}
%  \usepackage{algorithm}
%  \usepackage{algorithmic}
%  \usepackage{mdframed}
%  \usepackage{hyperref}
%  \usepackage{xcolor}
%  \hypersetup{
%  	colorlinks,
%  	linkcolor={red!40!gray},
%  	citecolor={blue!40!gray},
%  	urlcolor={blue!70!gray}
%  }
%  \usepackage{cleveref}

@@

% \begin{abstract}

\chapter{Exact expressions for double descent in determinantal random designs}
\label{ch:double_descent}

Building on the theory established in \cref{ch:design}, in this chapter
we analyze a determinantal ``surrogate'' random design to develop new
non-asymptotic theory on double descent in overparameterized linear models.
Double descent refers to the phase transition that is exhibited by
the generalization error of unregularized learning models when varying the ratio
between the number of parameters and the number of training
samples. The recent success of highly over-parameterized machine learning
models such as deep neural networks has motivated a theoretical analysis of
the double descent phenomenon in classical models such as linear
regression which can also generalize well in the over-parameterized
regime. We provide the first exact non-asymptotic
expressions for double descent of the minimum norm linear
estimator. Our approach involves constructing a special
determinantal point process  which we call surrogate random
design, to replace the standard i.i.d.~design of the training
sample. This surrogate design admits exact expressions for the mean
squared error of the estimator while preserving the key properties
of the standard design. We also establish an exact implicit
regularization result for over-parameterized training samples. In
particular, we show that, for the surrogate design, the implicit bias
of the unregularized minimum norm estimator precisely corresponds to
solving a ridge-regularized least squares problem on the population
distribution. In our analysis we introduce a new mathematical tool of
independent interest: the class of random matrices for which
determinant commutes with expectation.
Some of the results presented in this chapter were first published
in \fullcite{surrogate-design}.

% \end{abstract}

@@

%  To
% obtain our precise results, we use a somewhat non-standard random design, which
% we term surrogate random design (see Section~\ref{s:determinantal} for a
% detailed discussion), and which we expect will be of more general interest.
% Informally, the goal of a surrogate random design is to modify an original
% design to capture its main properties while being ``nicer'' in some useful way.
% In Theorem~\ref{t:asymptotic} and \Cref{sec:asymp-conj-details} we show, both
% theoretically and empirically, that our surrogate design accurately preserves
% the key properties of the original design when the data distribution is a
% multivariate Gaussian.

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}{0.49\textwidth}%
    \subcaption{Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
      numerical estimates even for non-isotropic
      features. Eigenvalue decay leads to a steeper
      descent curve in the under-determined regime ($n<d$).}
    \includegraphics[width=\textwidth]{design/figs/descent-intro}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.49\textwidth}
    \subcaption{The mean of the estimator $\X^\dagger\y$ exhibits
      shrinkage which closely matches the shrinkage of a
      ridge-regularized least squares optimum (theory lines), as characterized by
      Theorem \ref{t:unbiased}.}
    \includegraphics[width=\textwidth]{design/figs/descent-shrinkage}
  \end{subfigure}
  \caption{Illustration of the main results for $d=100$ and
    $\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
    eigenvalues decaying exponentially and scaled so that
    $\tr(\Sigmab^{-1})=d$. We use our surrogate
    formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
    \ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
    lines), accompanied by the empirical estimates based on the standard
    i.i.d.~design (error bars are three times the standard error of the
    mean). We consider three different condition numbers $\kappa$ of
    $\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
    i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
    $\w^*=\frac1{\sqrt{d}}\one$.}
  %\vspace{-5mm}
  \label{f:intro}
\end{figure}

Proof of Theorem \ref{t:mse} is given in \Cref{a:mse-proof}.
For illustration, we plot the MSE expressions in Figure~\ref{f:intro}a,
comparing them with empirical estimates of the true MSE under the
i.i.d.~design for a multivariate Gaussian distribution
$\mu=\Nc(\zero,\Sigmab)$ with several different covariance matrices $\Sigmab$. We keep the number of features $d$ fixed to
$100$ and vary the number of samples $n$, observing a double descent
peak at $n=d$. We observe that our theory aligns well with
the empirical estimates, whereas
previously, no such theory was available except for special
cases such as $\Sigmab=\I$ (more details in Theorem~\ref{t:asymptotic}
and Section~\ref{sec:asymp-conj-details}). The plots
show that varying the spectral decay of $\Sigmab$ has a significant effect on the
shape of the curve in the under-determined regime. We use the
horizontal line to denote the MSE of the null estimator
$\mathrm{MSE}[\zero]=\|\w^*\|^2=1$. When the eigenvalues of $\Sigmab$
decay rapidly, then the Moore-Penrose estimator suffers less error
than the null estimator for some values of $n<d$, and the curve
exhibits a local optimum in this regime.

One important aspect of Theorem~\ref{t:mse} comes from the relationship between $n$ and the parameter $\lambda_n$, which together satisfy $n=\tr(\Sigmab_\mu (\Sigmab_\mu+\lambda_n\I)^{-1})$.
This expression is precisely the classical notion of \emph{effective
  dimension} for ridge regression regularized with
$\lambda_n$~\citep{ridge-leverage-scores}, and it arises here even though there is
no explicit ridge regularization in the problem being considered in
Theorem~\ref{t:mse}.
The global solution to the ridge regression task (i.e., $\ell_2$-regularized
least squares) with parameter $\lambda$ is defined as:
\begin{align*}
  \argmin_\w \Big\{\E_{\mu,y}\big[\big(\x^\top\w-y(\x)\big)^2\big]
  + \lambda\|\w\|^2\Big\}\ =\ (\Sigmab_\mu +
  \lambda\I)^{-1}\v_{\mu,y},\quad\text{where } \v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x ].
\end{align*}
When Assumption \ref{a:linear} holds, then
$\v_{\mu,y}=\Sigmab_\mu\w^*$, however ridge-regularized least squares
is well-defined for much more general response models.
Our second result makes a direct connection between the (expectation
of the) unregularized minimum norm solution on the sample
and the global ridge-regularized solution.
While the under-determined regime (i.e., $n<d$) is of primary interest to us,
for completeness we state this result for arbitrary values of $n$ and $d$.
Note that, just like the definition of regularized least squares, this
theorem applies more generally than Theorem~\ref{t:mse}, in that it
does \emph{not} require the responses to follow any linear model as in
Assumption~\ref{a:linear} (proof in \Cref{s:unbiased-proof}).
\begin{theorem}[Implicit regularization of Moore-Penrose estimator]
  \label{t:unbiased}
  For $\mu$ satisfying
  % \footnote{The proof of Theorem \ref{t:unbiased}
  %   can be easily extended to probability measures $\mu$ that do not
  %   satisfy Assumption~\ref{a:general-position} (such as discrete distributions). We include this
  %   assumption here to simplify the presentation.}
  Assumption~\ref{a:general-position} and
  $y(\cdot)$ s.t.~$\v_{\mu,y}=\E_{\mu,y}[y(\x)\,\x]$ is well-defined,
  $\Xb\sim S_\mu^n$ (Definition \ref{d:surrogate}) and $\yb_i=y(\xbb_i)$,
  \begin{align*}
    \E\big[\Xb^\dagger\ybb\big] =
    \begin{cases}
      (\Sigmab_\mu + \lambda_n\I)^{-1}\v_{\mu,y} & \text{for }n<d,     \\
      \Sigmab_\mu^{-1}\v_{\mu,y}                 & \text{for }n \ge d,
    \end{cases}
  \end{align*}
  where, as in Theorem \ref{t:mse}, $\lambda_n$ is such that the effective dimension
  $\tr(\Sigmab_\mu(\Sigmab_\mu+\lambda_n\I)^{-1})$ equals $n$.
\end{theorem}

@@

\section{Related work}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.5\textwidth]{design/figs/descent-model-log}
  \caption{
    Surrogate MSE as a function of $d/n$, with $n$
    fixed to $100$ and varying dimension $d$ and condition number
    $\kappa$, for signal-to-noise ratio $\mathrm{SNR}=1$.}
  \label{f:model}
\end{figure}

@@

We first record an important property of the design $S_\mu^d$
which can be used to construct an over-determined design for any $n>d$. A similar
version of this result was also previously shown by
\cite{correcting-bias-journal} for a different determinantal design.

\begin{lemma}\label{l:decomposition}
  Let $\Xb\sim S_\mu^d$ and $\X\sim \mu^K$, where
  $K\sim\Poisson(\gamma)$. Then the matrix composed of a random
  permutation of the rows from $\Xb$ and $\X$ is distributed according to
  $S_\mu^{d+\gamma}$.
\end{lemma}

\begin{proof}
  Let $\Xt$ denote the matrix constructed from the permuted rows of
  $\Xb$ and $\X$.  Letting $\Z\sim\mu^{K+d}$, we derive the probability
  $\Pr\big\{\Xt\!\in\! E\big\}$ by summing over the possible index subsets  $S\subseteq
    [K+d]$ that correspond to the rows coming from $\Xb$:
  \begin{align*}
    \Pr\big\{\Xt\in E\big\} & = \E\bigg[\frac{1}{\binom{K+d}{d}}
      \sum_{S:\,|S|=d}\frac{\E[\det(\Z_{S,*})^2\one_{[\Z\in E]}\mid
    K]}{d!\det(\Sigmab_\mu)}\bigg]                               \\
                            & =\sum_{k=0}^\infty
    \frac{\gamma^k\ee^{-\gamma}}{k!}\,\frac{\gamma^dk!}{(k+d)!}\,
    \frac{\E\big[\sum_{S:\,|S|=d}\det(\Z_{S,*})^2\one_{[\Z\in E]}\mid
    K=k\big]}{\det(\gamma\Sigmab_\mu)}                           \\
                            & \overset{(*)}{=} \sum_{k=0}^\infty
    \frac{\gamma^{k+d}\ee^{-\gamma}}{(k+d)!}
    \,\frac{\E[\det(\Z^\top\Z)\one_{[\Z\in E]}\mid K=k]}{\det(\gamma\Sigmab_\mu)},
  \end{align*}
  where $(*)$ uses the Cauchy-Binet formula to sum over all subsets $S$
  of size $d$. Finally, since the sum shifts from $k$
  to $k+d$, the last expression can be rewritten as
  $\E[\det(\X^\top\X)\one_{[\X\in E]}]/\det(\gamma\Sigmab_\mu)$, where recall that
  $\X\sim\mu^K$ and $K\sim\Poisson(\gamma)$, matching the definition of $S_\mu^{d+\gamma}$.
\end{proof}

Another non-trivial property of the surrogate design $S_\mu^n$ is
that the expected sample size is in fact always equal to $n$.
\begin{lemma} \label{l:expected_size_n}
  Let $\Xb\sim S_\mu^n$ for any $n>0$.
  Then, we have $\E[\#(\Xb)] = n$.
\end{lemma}

\begin{proofof}{Lemma}{\ref{l:expected_size_n}}
  The result is obvious when $n=d$, whereas
  for $n>d$ it is an immediate consequence
  of Lemma \ref{l:decomposition}.
  Finally, for $n<d$ the expected sample
  size follows as a corollary of Lemma \ref{l:proj}, which states that
  \begin{align*}
    \text{(Lemma \ref{l:proj})} \qquad\E\big[\I - \Xb^\dagger\Xb\big] =
    (\gamma_n\Sigmab_\mu + \I)^{-1},
  \end{align*}
  where $\Xb^\dagger\Xb$ is the orthogonal projection onto
  the subspace spanned by the rows of $\Xb$. Since the rank of this
  subspace is equal to the number of the rows, we have
  $\#(\Xb)=\tr(\Xb^\dagger\Xb)$, so
  \begin{align*}
    \E\big[\#(\Xb)\big] = d - \tr\big((\gamma_n\Sigmab_\mu +
    \I)^{-1}\big) =
    \tr\big(\gamma_n\Sigmab_\mu(\gamma_n\Sigmab_\mu+\I)^{-1}\big) = n,
  \end{align*}
  which completes the proof.
\end{proofof}

Our general template for computing expectations under
a surrogate design $\Xb\sim\S_\mu^n$ is to use the following expressions based on the
i.i.d.~random design $\X\sim\mu^K$:
\begin{align}
  \E[F(\Xb)] & =\begin{cases}
                  \frac{\E[\det(\X\X^\top)F(\X)]}{\E[\det(\X\X^\top)]}
                  \quad K\sim\Poisson(\gamma_n) & \text{for }n<d, \\[2mm]
                  \frac{\E[\det(\X)^2F(\X)]}{\E[\det(\X)^2]}\hspace{3mm}
                  \quad K=d                     & \text{for }n=d, \\[2mm]
                  \frac{\E[\det(\X^\top\X)F(\X)]}{\E[\det(\X^\top\X)]}
                  \quad K\sim\Poisson(\gamma_n) & \text{for }n>d.
                \end{cases}\label{eq:cases}
\end{align}
These formulas follow from Definitions \ref{d:det} and
\ref{d:surrogate} because the determinants $\det(\X\X^\top)$ and
$\det(\X^\top\X)$ are non-zero precisely in the regimes $n\leq d$ and
$n\geq d$, respectively, which is why we can drop the restrictions on the
range of the Poisson distribution.
We compute the normalization constants
by introducing the concept of determinant preserving random matrices,
discussed in Section \ref{s:dp}.
% can be obtained using the following formulas: if $\X\sim\mu^K$ then
% \begin{align*}
% \text{(Lemma~\ref{l:normalization})}&&
% \E\big[\det(\X\X^\top)\big]  &= \ee^{-\gamma_n}\det(\I +
% \gamma_n\Sigmab_\mu)&&\text{for } K\sim\Poisson(\gamma_n),\ n<d,\\
%   \text{(Lemma~\ref{l:cb})}&& \E\big[\det(\X)^2\big] &=
%   d!\det(\Sigmab_\mu),&&\text{for } K=n=d,\\
%   \text{(Lemma~\ref{l:poisson})}&& \E\big[\det(\X^\top\X)\big]
% &=\det(\gamma_n\Sigmab_\mu),&&  \text{for }K\sim\Poisson(\gamma_n),\ n>d.
% \end{align*}
% \begin{remark}
% We will use $Z_\mu^n$ as a shorthand for the above normalization constants.
% \end{remark}
% For the case of $n=d$, the normalization constant can be found in the
% literature \cite{expected-generalized-variance,correcting-bias}:
% We prove Lemmas \ref{l:poisson} and \ref{l:normalization} in Section
% \ref{s:dp} by introducing the
% concept of determinant preserving random matrices.
% The lemmas play a
% crucial role in deriving a number of new expectation formulas for the
% under- and over-determined surrogate designs that we use to prove Theorems~\ref{t:mse}
% and~\ref{t:unbiased} in Section~\ref{s:expectations}.
% On the other hand, Lemma  \ref{l:cb} and the design $S_\mu^d$ can be found in the
% literature \citep{expected-generalized-variance,correcting-bias-journal},
% and we will rely on those known results in this case.
% Importantly, the $n=d$ case offers a continuous transition between the under-
% and over-determined regimes because the distribution $S_\mu^n$
% converges to $S_\mu^d$ when $n$ approaches $d$ from above and
% below.

\paragraph{Proof sketch of Theorem \ref{t:mse}}
% We now highlight some key expectation formulas for surrogate designs, which are used to
% derive the MSE expressions from Theorem \ref{t:mse}. 
We focus here on the under-determined regime (i.e., $n<d$),
highlighting the key new expectation formulas we develop to derive the
MSE expressions for surrogate designs. A standard decomposition of the MSE yields:
\begin{align}
  \MSE{\Xb^\dagger\ybb}
   & = \E\big[\|\Xb^\dagger(\Xb\w^*+\xib)-\w^*\|^2\big]
  =\sigma^2\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big] +
  \w^{*\top}\E\big[\I-\Xb^\dagger\Xb\big]\w^*.\label{eq:mse-derivation}
\end{align}
Thus, our task is to find closed form expressions for the two
expectations above. The latter, which is the expected projection onto
the complement of the row-span of $\Xb$, is proven in
\Cref{s:unbiased-proof}.
\begin{lemma}\label{l:proj}
  If  $\Xb\sim S_\mu^n$ and $n<d$, then we have:
  $\E\big[\I-\Xb^\dagger\Xb\big] = (\gamma_n\Sigmab_\mu+\I)^{-1}$.
\end{lemma}

\noindent
No such expectation formula is known for i.i.d.~designs, except when
$\mu$ is an isotropic Gaussian. In \Cref{s:unbiased-proof}, we
also prove a generalization of Lemma \ref{l:proj} which is then used
to establish our implicit regularization result
(Theorem~\ref{t:unbiased}). We next give an expectation formula
for the trace of the Moore-Penrose inverse of the covariance
matrix for a surrogate design (proof in \Cref{a:mse-proof}).
\begin{lemma}\label{l:sqinv-all}
  If  $\Xb\sim S_\mu^n$ and $n<d$, then:
  $\E\big[\tr\big((\Xb^\top\Xb)^{\dagger}\big)\big]
    =\gamma_n\big(1-
    \det\!\big((\tfrac1{\gamma_n}\I+\Sigmab_\mu)^{-1}\Sigmab_\mu\big)\big)$.
\end{lemma}

@@

We next give a few simple examples to provide some intuition. First, note
that every $1\times 1$ random matrix is determinant preserving simply
because taking a determinant is an identity transfomation in one
dimension. Similarly, every fixed matrix is determinant preserving because
in this case taking the expectation is an identity
transformation. In all other cases, however, Definition \ref{d:main}
has to be verified more carefully. Further examples (positive and
negative) follow.
\begin{example}
  If $\A$ has i.i.d. Gaussian entries $a_{ij}\sim\Nc(0,1)$, then
  $\A$ is d.p.~because $\E[\det(\A)]=0$.
\end{example}

In fact, it can be shown that all random matrices with independent entries
are determinant preserving. However, this is not a necessary condition.
\begin{example}\label{e:rank-1}
  Let $\A = s\,\Z$, where $\Z$ is fixed with $\rank(\Z) = r$, and $s$
  is a scalar random variable. Then for $|\Ic|=|\Jc|=r$ we have
  \begin{align*}
    \E\big[\det(s\,\Z_{\Ic,\Jc})\big] & = \E[s^r]\det(\Z_{\Ic,\Jc})
    =\det\Big(\big(\E[s^r]\big)^{\frac1r}\,\Z_{\Ic,\Jc}\Big),
  \end{align*}
  so if $r=1$ then $\A$ is determinant preserving, whereas if $r>1$
  and $\Var[s]>0$ then it is not.
\end{example}
